%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Thesis 
% LaTeX Template
% Version 1.2 (29/7/12)
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original authors:
% Steven Gunn 
% http://users.ecs.soton.ac.uk/srg/softwaretools/document/templates/
% and
% Sunil Patel
% http://www.sunilpatel.co.uk/thesis-template/
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
% Note:
% Make sure to edit document variables in the Thesis.cls file
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-----------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%-----------------------------------------------------------------------------

\documentclass[11pt,a4paper, wide, twoside]{IIBproject}

\graphicspath{{./Pictures/}} % Specifies the directory where pictures are stored

\usepackage[round, sort&compress]{natbib} % Use the natbib 
  %reference package - read up on this to edit the reference style; 
  %if you want text (e.g. Smith et al., 2012) for the in-text references 
  %(instead of numbers), remove 'numbers' 

\usepackage[usenames,dvipsnames]{xcolor}

\usepackage{caption}
\captionsetup{margin={15pt},parskip=10pt,format=hang,indention=-.8cm}

\usepackage{type1cm}
\usepackage{minitoc}
\usepackage{lettrine}
\usepackage{mathtools}
\usepackage[version=3]{mhchem}

\usepackage{tikz}
\usetikzlibrary{matrix,shapes,arrows,patterns,positioning}

\tikzstyle{block} = [draw, fill=blue!20, rectangle, minimum height=3em, 
                      minimum width=6em]
\tikzstyle{sum} = [draw, fill=blue!20, circle, node distance=1cm]
\tikzstyle{input} = [coordinate]
\tikzstyle{output} = [coordinate]
\tikzstyle{branch}=[fill,shape=circle,minimum size=3pt,inner sep=0pt]
\tikzstyle{label} = [text width=7em, text centered, font=\itshape]

\tikzstyle{key} = [draw, circle, fill=black, minimum width=5mm, inner sep=0pt,
  minimum height=5mm, text centered, text=white,font=\small\bfseries]

%for HMM diagram
\tikzstyle{match} = [draw, fill=blue!20, rectangle, inner sep=0pt, 
                    minimum height=1cm, minimum width=1cm, font=\footnotesize]
\tikzstyle{insert} = [draw, fill=yellow!40, diamond, inner sep=0pt,  
                    minimum height=1cm, minimum width=1cm, font=\footnotesize]
\tikzstyle{delete} = [draw, fill=red!20, circle, inner sep=0pt,  
                    minimum height=1cm, minimum width=1cm, font=\footnotesize]

\tikzstyle{ppr} = [draw=black, circle, minimum width=3mm, inner sep=0pt,
  minimum height=3mm, text centered, font=\tiny\bf, outer sep=0.5mm]
\tikzstyle{induce} = [->, >=triangle 60]
\tikzstyle{repress} = [->, >=|]

\usepackage{pgfplots}
\pgfplotsset{compat=1.7}

\usepackage{subcaption}


\begin{document}

\setstretch{1.5}

\pagenumbering{roman}
% PDF meta-data
%\hypersetup{pdftitle={\ttitle}}
%\hypersetup{pdfsubject=\subjectname}
%\hypersetup{pdfauthor=\authornames}
%\hypersetup{pdfkeywords=\keywordnames}

%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

% Title Page
\author{Haydn \textsc{King} (JE)}
\name{Haydn King}
\title{A Cross-Genome Study of the Pentatricopeptide Repeat Protein}
\projectgroup{F}
\maketitle

%----------------------------------------------------------------------------------------
%	ABSTRACT PAGE
%----------------------------------------------------------------------------------------
\setcounter{page}{1}

\abstract{

Synthetic biology is a new and developing engineering field which attempts to
apply proven engineering concepts and techniques such as standardisation,
characterisation and encapsulation to biology in order to design new and useful
forms of life.

The field often makes use of analogies with electrical engineering -- in
synthetic biology, well characterised components are combined and 
interact together in biological circuits.
Circuits have been designed to perform a wide variety of tasks, such as
the production of useful chemicals, detecting the presence of harmful toxins
or producing bioluminescent light.

Since designing novel proteins from scratch is beyond our technological reach 
at present, progress in synthetic biology usually results from the
characterisation of natural systems found in one organism which are then
applied in different situations.

This project investigates such a system commonly found in plant cells, where
the cell nucleus is thought to control gene expression in the chloroplasts and
other organelles using signals encoded by members of the Pentatricopeptide
Repeat (PPR) protein family.
This system is particularly interesting for synthetic biology as there is a
deficit of well characterised parts for accurately controlling the production
of an arbitrary protein.
PPRs have been shown to be able to do this naturally -- by binding with 
targeted mRNA molecules in the chloroplast, they are able to either increase of
decrease expression levels of particular proteins.
It is hoped that characterisation of this class of proteins will allow the
design of PPRs which can control expression of an arbitrary protein.

In addition to their development as a tool for synthetic biology, an
understanding of the interactions between PPRs and mRNA will be vital to
understanding the control mechanisms at work in the chloroplast, which is a
necessary step if we wish to exploit the large potential for biosynthesis
within the chloroplast.

The first part of the project is concerned with automating the detection and
annotation of naturally occurring PPRs in existing genomes.
While PPRs have been discovered in a wide variety of plants, a specific, 
clearly defined and repeatable method for their detection and localisation
has not.
An algorithm for reliable and repeatable PPR discovery was developed in the
first part of the project.

In the second part, this algorithm was applied to a large variety of plant 
genomes in order to assess the abundance of PPR proteins.
It was found that the majority of the plants surveyed had a similar number of
PPRs (between 400 and 600).
It is hypothesised that this is due to a considerable conservation of function
of PPRs across genomes.

Each member of the PPR family binds to a particular RNA sequence, but
attempts to directly predict the RNA binding sequence of a given PPR 
were thwarted by the limited number of well characterised PPRs.
However, the binding sites of a set of characterised PPRs was identified in 
the model organism \emph{A. thaliana}.
Likely homologs of the genes containing these binding sites were then found in
the chloroplast genomes of numerous other plants and regions of very high
sequence similarity to the original binding sites were found, supporting the
hypothesis that PPR function is highly conserved across different species.

The final part of the project studied the usefulness of PPRs in the context of
synthetic biology.
If, given a target binding sequence, we could design a PPR protein which would
bind to this sequence then it was established that it should be possible to 
construct logic gates using known PPR-RNA interactions.

The fact that each each PPR would act as an isolated signalling modality has
the potential to revolutionise the field as in most projects problems due to
cross-talk of reporter molecules fundamentally limit the complexity of the
circuit.

}

\clearpage % Start a new page

%----------------------------------------------------------------------------------------
%	LIST OF CONTENTS/FIGURES/TABLES PAGES
%----------------------------------------------------------------------------------------

\setcounter{tocdepth}{1}
\tableofcontents % Write out the Table of Contents

\clearpage % Start a new page

%----------------------------------------------------------------------------------------
%	CONTENT - CHAPTERS
%----------------------------------------------------------------------------------------

\pagestyle{mainstyle}
\pagenumbering{arabic}

% Include the chapters of the thesis as separate files from the Chapters folder
% Uncomment the lines as you write the chapters

\input{./Chapters/Chapter1}
\input{./Chapters/Chapter2} 
\input{./Chapters/Chapter3}
\input{./Chapters/Chapter4} 
\input{./Chapters/Chapter5} 
\input{./Chapters/Chapter6} 
%\input{./Chapters/Chapter7} 

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - APPENDICES
%----------------------------------------------------------------------------------------

\addtocontents{toc}{\vspace{1em}} % Add a gap in the Contents, for aesthetics

\appendix % Cue to tell LaTeX that the following 'chapters' are Appendices

\renewcommand{\chaptermark}[1]{ \markboth{Appendix \thechapter.\ #1}{} }
% Include the appendices of the thesis as separate files from the Appendices folder
% Uncomment the lines as you write the Appendices

\input{./Appendices/AppendixA}
\input{./Appendices/AppendixB}
%\input{./Appendices/AppendixC}

%\addtocontents{toc}{\vspace{1em}} % Add a gap in the Contents, for aesthetics


%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

\label{Bibliography}

\bibliographystyle{abbrvnat} % Use the "unsrtnat" BibTeX style for formatting the Bibliography

\bibliography{Bibliography} % The references (bibliography) information are stored in the file named "Bibliography.bib"

\end{document}  
