
\chapter{Conclusions and Further Work}
\label{chap:Conclusions} 

\lettrine{A}{} brief summary of the work carried out during the project and a 
description of the work required to make PPR technology a reality are given in 
this chapter.

\section{Summary of the Work}

Pentatricopeptide repeat (PPR) proteins are a vital class of proteins for a 
large number of plants.
During this project, software was developed to automatically identify, extract 
and characterise PPRs from unannotated DNA sequence.
It was found that the majority of plants for which nuclear genomes are 
available contain roughly the same number of PPR proteins, suggesting that the
function of these proteins is consistent across these plants
\footnote{All extracted PPRs are available from
\emph{https://github.com/haydnKing/4th-year-project-data/tree/master/PPRs}}.

A comparison of current theories regarding the prediction of PPR binding
preferences was made, and it was shown that while current models of PPR binding
sites score a genuine match highly, they allow too much variation in the
sequence such that any reasonably long random sequence will contain matches 
which are just as good if not better, making accurate prediction impossible.

It was also shown that there appears to be considerable conservation of PPR 
binding sites across different organisms.
If this is confirmed experimentally then there must also be considerable 
conservation of both PPR function and PPR binding preferences across the same 
organisms.
This conservation would make experimentally extracting and characterising new
proteins from these organisms relatively simple, which would drastically 
increase the number of PPR-RNA pairs of which we are aware.

In order to assess the potential of the PPR protein as a tool for the wider
field of synthetic biology, a general model of PPR activity based on known 
interactions was created.
By simulating various interaction networks, it was shown that PPRs can be 
used to perform various simple logic operations at the translation level within
the cell.
Since each PPR can be programmed to recognise long sequences of RNA (up to at
least $30$bp) this leads to an enormous address space for signalling (a maximum
of $4^{30}$), effectively solving the problem of cross-talk in molecular 
circuit design.

\section{Future Work and Directions}

Increasing the number of known PPR-RNA will allow us to improve the prediction
of how binding specificity is encoded, and will eventually enable us to design
PPRs which can bind to any sequence of appropriate length of our choice.

However, a thorough understanding of the rules which govern binding domain 
preference is
only part of the way to realising the potential of PPR-based technology.
We must also understand the molecular basis behind the interactions with
translation and mRNA stability in order to exploit these effects for our own
goals.

Although plausible models exist suggesting RNA stabilisation, ribosome 
recruitment and degradation of mRNA, and there is evidence for some of 
these interactions
\emph{in vitro} \citep[see][]{Prikryl2011}, there is little characterisation of
these mechanisms \emph{in vivo}, nor of how PPRs should
be designed to exploit these mechanisms.

